### 2017-12-20 paper/response_to_reviewers.tex

\setlength\parindent{0pt}

\begin{document}

\textbf{RE: MEE-17-08-729 }\\
\bigskip
Dear Dr Vamosi,\\
\bigskip

I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
I have taken all of their comments on board, and respond to their points in details below.
For improving clarity in this document, reviewers comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
Also, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.

@@

\item{\textcolor{blue}{Secondly, problematic implementations.
Here there is again only one major issue, and that concerns the time-slicing options.
Initially it should be noted by the author how useful this feature could be in generating time series of disparity that are long enough for time series statistics to be applied.
Currently most palaeobiologists are stuck using binned data (e.g., Figure 4) which tends to lead to time series of insufficient length (here N = 3) to explore, for example, correlations with climate proxies (or some other hypothesised driver of morphological diversity).
Time-slicing could revolutionise the field by opening up data sets to such analyses and an example alongside the binning in Figure 4 should be considered.
However, the current implemented options are, in my view, deeply problematic.
Specifically, dispRity offers four options (L145-156), but in effect these all boil down to treating a point along a branch as either identical to the descendant (end of branch) or the ancestor (start of branch).
However, I think most biologists would treat such a value as more likely to be intermediate between these extremes and possessing some form of uncertainty depending on the generating model's assumptions (e.g., Brownian motion).
I don't mind offering a punctuated model (descendant value propagated back along branch) but some form of intermediate value should also exist.
Without this option data are likely to be biased, e.g., towards generating more "jagged" time series and perhaps more significant slice-to-slice changes (using the packages "test" functions) than are realistic (ie.e, type I errors).}}

The standard approach in palaeobiology relies on the following pipeline: (i) coding characters for tips; (ii) running ancestral states reconstructions for the nodes; (iii) ordinating the resulting matrix (usually via a PCO); (iv) measuring disparity from there (see Lloyd 2016 \textit{Biol J Lin Soc} for an excellent description of this pipeline).
I feel that the reviewer's suggestion of proposing more accurate models for ancestral states reconstructions and will technically not be doable after step (iii).
In fact, the ordination, to be meaningful as a ``morphospace'' should always be done on the full dataset.
I.e. no additional data should be added to the space \textit{a posteriori}.
In practice, the \textit{time.subset} function can only select between two information: the ancestral node (DELTRAN) and the descendant node/tip (ACCTRAN) along the edge since only the two of them are real objects in the morphospace.
It is therefore not possible to create an ancestral states reconstruction along the edge after running the ordination.

I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges
I have modified the name of the two previous ``punctuated'' and ``gradual'' algorithms to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g. if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).

@@

I've added these models descriptions in the manual and in the manuscript:

\begin{itemize}
    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\textit{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).} lines @@@

### 2017-12-20 paper/dispRity_paper_highlighted.tex
Minor fixes

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these \hl{to} a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the multidimensional volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.

@@

% Multidimensionality
Biological data are complex.
To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable in our physical world restricted to three dimensions.
These multivariate analyses can be used to investigate changes in morphological diversity through time \citep{Close2015}, competitive replacement scenarios \citep{Brusatte12092008}, relationships among form and function \citep{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep{raup1966geometric}.
The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g. \citealt{brusattedinosaur2012}; and continuous traits such as lengths, e.g. \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
\hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
\citealt{lloyd2016estimating}
\hl{and references therein; e.g. in}
\citealt{Close2015}
\hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
\citealt{PCA}
\hl{; e.g. in}
\citealt{zelditch2012geometric}
\hl{), a metric scaling (PCO, PCoA,}
\citealt{PCO}
\hl{; e.g. in}
\citealt{Brusatte12092008}
\hl{) or a non-metric scaling (MDS, NMDS,}
\citealt{MDS}
\hl{; e.g. in}
\citealt{Liow2004,DonohueDim}
\hl{).}
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.

@@

Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
\hl{NPMANOVA, }
\citealt{NPMANOVA}
\hl{; e.g. in }
\citealt{Brusatte12092008}
\hl{)}
multidimensional permutation tests
\hl{(}
\citealt{ManlyPermutations}
\hl{; e.g. in }
\citealt{diaz2016global}
\hl{)}
or even simply by looking at the confidence interval overlaps between disparity measurements
\hl{().}
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

% Problems with this variety
In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric},
\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
\hl{as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)}
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
\hl{This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).}
This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!

% Here comes the solution!
The aim of the \disp package is to solve such problems by providing a completely flexible framework for studying multidimensional data.
This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).

\section{Description}
In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}
(see Manuals and Vignettes below).
Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).% and Table \ref{Tab:main_fun}).
\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g. the multidimensional space can be a distance matrix][]{Close2015}.

@@

% \begin{table}
%     \begin{tabular}{ll}
%         \hline
%         Function & Description \\ 
%         \hline
%         \texttt{geomorph.ordination} & Imports \hl{shape} data from the \texttt{geomorph} package. \\
%         \texttt{Claddis.ordination} & Imports \hl{discrete morphological} data from the \texttt{Claddis} package. \\
%         \texttt{custom.subsets} & Separates data into custom subsets. \\
%         \texttt{time.subsets} & Separates data into time subsets. \\
%         \texttt{boot.matrix} & Bootstraps and rarefies a matrix or a \disp object. \\
%         \disp & Calculates disparity from an matrix or a \disp object. \\
%         \texttt{summary.dispRity} & Summarises a \disp object. \\
%         \texttt{plot.dispRity} & Plots a \disp object. \\
%         \texttt{test.dispRity} & Applies statistical tests to a \disp object.\\
%         \texttt{dispRity.per.group} & Pipeline for \hl{disparity-within-groups} analysis. \\
%         \texttt{dispRity.through.time} & Pipeline for disparity-through-time analysis. \\
%         \hline
%     \end{tabular}
%     \caption{Main functions in the \disp package.}
%     \label{Tab:main_fun}
% \end{table}

@@

\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).
\label{Tab:metrics}

@@

\hl{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).}

@@

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
These results can then be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
% as common arguments which central tendency (e.g. if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base R.
% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of subsets to compare.

@@

\subsection{Typical disparity among groups analysis}
One typical question with such analysis would be to test whether two groups of species have a different disparity.
For example, using the data described above, we can test whether the crown mammals are more diverse in term of morphology than the stem ones.
In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.

\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that selects :

\noindent \texttt{> mammal\_groups <- crown.stem(BeckLee\_tree)}

@@

\noindent \texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}

@@

\noindent \texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

@@

\noindent \texttt{> disparity}\\
\noindent \texttt{ ---- dispRity object ---- }\\
\noindent \texttt{2 customised subsets for 50 elements with 48 dimensions:}\\
          \texttt{crown, stem.}\\
\noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
\noindent \texttt{Disparity was calculated as: c(sum, variances).}\\

@@

\noindent \texttt{> summary(disparity)}

\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrrr}
  \hline
 & subsets & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
  \hline
1 & crown &  30 & 2.00 & 1.93 & 1.87 & 1.92 & 1.95 & 1.98 \\ 
  2 & stem &  20 & 1.72 & 1.63 & 1.53 & 1.60 & 1.66 & 1.69 \\ 
   \hline
\end{tabular}
\caption{Summarising a \disp object (disparity per groups). $n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and 95\% are the confidence intervals.}
\label{Tab:summary_group}
\end{table}

\noindent \texttt{> plot(disparity)}

@@

\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}\\

@@

\subsubsection{Typical disparity-through-time analysis}
A 
\hl{subsequent} 
question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).

@@

\noindent \texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}

@@

\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
\noindent \texttt{> \hl{bootstraps2} <- boot.matrix(\hl{time\_subsets})}\\
\noindent \texttt{> \hl{disparity2} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\

@@

\noindent \texttt{> summary(\hl{disparity2})}

@@

\noindent \texttt{> plot(\hl{disparity2}, type = \textquotedbl continuous\textquotedbl)}

@@

\noindent \texttt{> test.dispRity(disparity2, test = adonis.dispRity)}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \hline
              & Df & Sum Sq & Mean Sq & F Model & R2   & Pr($>$F) \\ 
  \hline
    time      & 2  &  7.50 & 3.75     & 2.06    & 0.11 & <0.01 *** \\ 
    Residuals & 34 & 61.82 & 1.81     &         & 0.89 & \\ 
    Total     & 36 & 69.32 &          &         & 1.00 & \\ 
   \hline
\end{tabular}
\caption{Permanova output from \texttt{adonis.dispRity}:\\
Call: vegan::adonis(formula = dist(matrix) \texttildelow time, data = disparity2, method = "euclidean"); Permutation: free; Number of permutations: 999; Terms added sequentially (first to last). Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1 ` ' 1.}
\label{Tab:anova}
\end{table}

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.

>>>>

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these \hl{to} a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space}.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure \hl{the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.

@@

% Multidimensionality
Biological data are complex.
To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g. \citealt{Close2015}; and continuous traits such as lengths, e.g. \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
\hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
\citealt{lloyd2016estimating}
\hl{and references therein; e.g. in}
\citealt{Close2015}
\hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
\citealt{PCA}
\hl{; e.g. in}
\citealt{zelditch2012geometric}
\hl{), a metric scaling (PCO, PCoA,}
\citealt{PCO}
\hl{; e.g. in}
\citealt{Brusatte12092008}
\hl{) or a non-metric scaling (MDS, NMDS,}
\citealt{MDS}
\hl{; e.g. in}
\citealt{Liow2004,DonohueDim}
\hl{).}
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.

@@

Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
\hl{NPMANOVA, }
\citealt{NPMANOVA}
\hl{; e.g. in }
\citealt{Brusatte12092008}
\hl{)}
multidimensional permutation tests
\hl{(}
\citealt{ManlyPermutations}
\hl{; e.g. in }
\citealt{diaz2016global}
\hl{)}
or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements
\hl{().}
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

% Problems with this variety
In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
\hl{as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)}
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
\hl{This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).}
This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!

% Here comes the solution!
The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
The package is described here for the use of discrete morphological data diversity analysis but can be generalised to any type of multidimensional data (see the package vignette for more examples).

\section{Description}
In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
\hl{calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation}.
Some additional functions modify the multidimensional space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
\hl{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements} \citep[e.g. the multidimensional space can be a distance matrix][]{Close2015}.

@@

@@

\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
\label{Tab:metrics}

@@

\hl{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type II error (thus decreasing accuracy).
See Guillerme \& Cooper \textit{in review} for full description.}

@@

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
% as common arguments which central tendency (e.g. if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.
% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets} - this can be used for \texttt{lm} class functions) or (5) any list of pairs (or more) of subsets to compare.

@@

\subsection{Example}

The \texttt{R} script for the following example is available in the supplementary materials.

\subsubsection{Typical disparity among groups analysis}
One typical question with such analysis would be to test whether two groups of species have a different disparity.
For example, using the data described above, we can test whether the crown mammals are more diverse in term of morphology than the stem ones.
In other words, whether the approximation of the volume within the morphospace (the disparity as defined above) is different in crown or stem mammals.

\hl{\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that separate the crown and stem species given a phylogeny (allowing to ignore the nodes or not):}

\noindent
\hl{\texttt{> mammal\_groups <- crown.stem(BeckLee\_tree, inc.nodes = FALSE)}}

@@

\noindent
\texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50,}\\
\texttt{group = mammal\_groups, metric = c(sum, variances))}

@@

\noindent
\texttt{> subsets <- custom.subsets(data = BeckLee\_mat50, group = mammal\_groups)}\\
\noindent
\texttt{> bootstraps <- boot.matrix(subsets)}\\
\noindent
\texttt{> \hl{disparity2} <- dispRity(bootstraps, metric = c(sum, variances))}\\

@@

\noindent \texttt{> \hl{disparity2}}\\
\noindent \texttt{ ---- dispRity object ---- }\\
\noindent \texttt{2 customised subsets for 50 elements with 48 dimensions:}\\
          \texttt{crown, stem.}\\
\noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
\noindent \texttt{Disparity was calculated as: c(sum, variances).}\\

@@

\noindent \texttt{> summary(\hl{disparity2)}}

\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrrr}
  \hline
 & subsets & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
  \hline
1 & crown &  30 & 2.00 & 1.93 & 1.87 & 1.92 & 1.95 & 1.98 \\ 
  2 & stem &  20 & 1.72 & 1.63 & 1.53 & 1.60 & 1.66 & 1.69 \\ 
   \hline
\end{tabular}
\caption{Summarising a \disp object (disparity per groups). $n$ is the number of elements per subsets, $obs$ the observed disparity (not bootstrapped), $bs.median$ is the median bootstrapped disparity (here the median of the sum of ranges) and the 2.5, 25, 75 and
\hl{97.5}
\% are the confidence intervals.}
\label{Tab:summary_group}
\end{table}

\noindent \texttt{> plot(\hl{disparity2)}}

@@

\noindent \texttt{> test.dispRity(\hl{disparity2}, test = wilcox.test)}\\

@@

\subsubsection{Typical disparity-through-time analysis}
A 
\hl{subsequent} 
question to this observation could be to test whether this difference is due to an overall change in disparity through time or not.
In other words, did disparity increase through time in mammals globally or is it only between stem and crown mammals.
Using the same definition of the multidimensional space and disparity as in the previous example, we can measure, for example, changes in disparity through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).
\hl{Note that stratigraphic times can be generated automatically using the \texttt{get.bin.ages} utility function.}

@@

\noindent
\texttt{> \hl{disparity3} <- dispRity.through.time(data = BeckLee\_mat50,}\\
\texttt{tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}

@@

\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree,}\\ \texttt{time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
\noindent \texttt{> \hl{bootstraps2} <- boot.matrix(\hl{time\_subsets})}\\
\noindent \texttt{> \hl{disparity4} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\

@@

\noindent \texttt{> summary(\hl{disparity4})}

@@

\noindent \texttt{> plot(\hl{disparity4}, type = \textquotedbl continuous\textquotedbl)}

@@

\noindent \texttt{> test.dispRity(\hl{disparity4, test = adonis.dispRity)}}

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \hline
              & Df & Sum Sq & Mean Sq & F Model & R2   & Pr($>$F) \\ 
  \hline
    time      & 2  &  7.50 & 3.75     & 2.06    & 0.11 & <0.01 *** \\ 
    Residuals & 34 & 61.82 & 1.81     &         & 0.89 & \\
    Total     & 36 & 69.32 &          &         & 1.00 & \\ 
   \hline
\end{tabular}
\caption{
\hl{Permanova output from \texttt{adonis.dispRity}:\\
Call: vegan::adonis(formula = dist(matrix) \texttildelow  time, data = disparity2, method = "euclidean"); Permutation: free; Number of permutations: 999; Terms added sequentially (first to last). Signif. codes:  0 `***' 0.001 `**' 0.01 `*' 0.05 `.' 0.1 ` ' 1.}
}
\label{Tab:anova}
\end{table}

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test}.
